---
title: 'ESM 244 Lab Week 8: Clustering (k-means & hierarchical)'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(palmerpenguins)
library(plotly)

# Packages for cluster analysis:
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```

## Intro to cluster analysis (k-means, hierarchical)

### Part 1. K-means clustering: 

Penguin clusters 

```{r}

ggplot(penguins) +
  geom_point(aes(x = bill_length_mm, 
                 y = bill_depth_mm, 
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c("orange","cyan4","darkmagenta"))

ggplot(penguins) +
  geom_point(aes(x = flipper_length_mm, 
                 y = body_mass_g, 
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c("orange","cyan4","darkmagenta"))

# How many clusters do you THINK there should be? 
number_est <- NbClust(penguins[3:6], min.nc = 2, max.nc = 10, method = "kmeans")

# By these estimators, 2 is the best number of clusters...but should that change our mind? Maybe...but here I think it makes sense to still stick with 3 and see how it does. 

# What if we consider similarities across all four variables? 
penguins_complete <- penguins %>% 
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm)

penguins_scale <- penguins_complete %>% 
  select(ends_with("mm"), body_mass_g) %>% 
  scale()

penguins_km <- kmeans(penguins_scale, 3) # kmeans specifying 2 groups to start

# See what it returns (different elements returned by kmeans function):
penguins_km$size
penguins_km$centers
penguins_km$cluster

# Bind the cluster number to the original data

penguins_cl <- data.frame(penguins_complete, cluster_no = factor(penguins_km$cluster))

ggplot(penguins_cl) +
  geom_point(aes(x = flipper_length_mm, 
                 y = body_mass_g, 
                 color = cluster_no,
                 shape = species))
```

Then get counts:
```{r}
penguins_cl %>% 
  count(species, cluster_no) %>% 
  pivot_wider(names_from = cluster_no, values_from = n) %>% 
  rename('Cluster 1' = '1', 'Cluster 2' = '2', 'Cluster 3' = '3')

```

Takeaway: *most* chinstraps in Cluster 1, and *most* Adelies in Cluster 2, and *all* Gentoos are in Cluster 3 by k-means clustering. 

### Part 2. Cluster analysis: hierarchical

Hierarchical cluster analysis (dendrograms) in R

Relevant functions:

stats::hclust() - agglomerative hierarchical clustering
cluster::diana() - divisive hierarchical clustering

We'll be using WorldBank environmental data (simplified), wb_env.csv
```{r}

# Get the data
wb_env <- read_csv("wb_env.csv")

# Only keep top 20 greenhouse gas emitters (for simplifying visualization here...)
wb_ghg_20 <- wb_env %>% 
  arrange(-ghg) %>% 
  head(20)

# Scale it (can consider this for k-means clustering, too...)
wb_scaled <- as.data.frame(scale(wb_ghg_20[3:7]))

# Update to add rownames (country name)
rownames(wb_scaled) <- wb_ghg_20$name

# Compute dissimilarity values (Euclidean distances):
diss <- dist(wb_scaled, method = "euclidean")

# Hierarchical clustering (complete linkage)
hc_complete <- hclust(diss, method = "complete" )

# Plot it (base plot):
plot(hc_complete, cex = 0.6, hang = -1)

```

Divisive clustering:
```{r}
hc_div <- diana(diss)

plot(hc_div, hang = -1)
rect.hclust(hc_div, k = 4, border = 2:5)
```

We might want to compare those...because they differ slightly.
```{r}

# Convert to class dendrogram
dend1 <- as.dendrogram(hc_complete)
dend2 <- as.dendrogram(hc_div)

# Combine into list
dend_list <- dendlist(dend1,dend2)

# Make a tanglegram
tanglegram(dend1, dend2)

# Convert to class 'dendro' for ggplotting
data1 <- dendro_data(hc_complete)

# Simple plot with ggdendrogram
ggdendrogram(hc_complete, 
             rotate = TRUE) +
  theme_minimal() +
  labs(x = "Country")

# Want to do it actually in ggplot? Here: 
label_data <- bind_cols(filter(segment(data1), x == xend & x%%1 == 0), label(data1))

ggplot() + 
geom_segment(data=segment(data1), aes(x=x, y=y, xend=xend, yend=yend)) +
geom_text(data=label_data, aes(x=xend, y=yend, label=label, hjust=0), size=2) +
coord_flip() + 
scale_y_reverse(expand=c(0.2, 0)) +
theme_bw() +
theme(panel.border = element_blank(),
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      axis.line = element_blank(),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "None") 

```


####Part 2. Cluster analysis: hierarchical

Hierarchical cluster analysis (dendrograms) in R

Relevant functions:

stats::hclust() - agglomerative hierarchical clustering
cluster::diana() - divisive hierarchical clustering

We'll be using WorldBank environmental data (simplified), wb_env.csv
```{r}

# Get the data
wb_env <- read_csv("wb_env.csv")

# Only keep top 20 greenhouse gas emitters (for simplifying visualization here...)
wb_ghg_20 <- wb_env %>% 
  arrange(-ghg) %>% 
  head(20)

# Scale it (can consider this for k-means clustering, too...)
wb_scaled <- as.data.frame(scale(wb_ghg_20[3:7]))

# Update to add rownames (country name)
rownames(wb_scaled) <- wb_ghg_20$name

# Compute dissimilarity values (Euclidean distances):
diss <- dist(wb_scaled, method = "euclidean")

# Hierarchical clustering (complete linkage)
hc_complete <- hclust(diss, method = "complete" )

# Plot it (base plot):
plot(hc_complete, cex = 0.6, hang = -1)

```

Divisive clustering:
```{r}
hc_div <- diana(diss)

plot(hc_div)
```

We might want to compare those...because they differ slightly.
```{r}

# Convert to class dendrogram
dend1 <- as.dendrogram(hc_complete)
dend2 <- as.dendrogram(hc_div)

# Combine into list
dend_list <- dendlist(dend1,dend2)

# Make a tanglegram
tanglegram(dend1, dend2)

# Convert to class 'dendro' for ggplotting
data1 <- dendro_data(hc_complete)

# Simple plot with ggdendrogram
ggdendrogram(hc_complete, 
             rotate = TRUE) +
  theme_minimal() +
  labs(x = "Country")

# Want to do it actually in ggplot? Here: 
label_data <- bind_cols(filter(segment(data1), x == xend & x%%1 == 0), label(data1))

ggplot() + 
geom_segment(data=segment(data1), aes(x=x, y=y, xend=xend, yend=yend)) +
geom_text(data=label_data, aes(x=xend, y=yend, label=label, hjust=0), size=2) +
coord_flip() + 
scale_y_reverse(expand=c(0.2, 0)) +
theme_bw() +
theme(panel.border = element_blank(),
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      axis.line = element_blank(),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "None") 

```


